Selection coefficients at the phylogenetic and population-genetic scale

### 2022-04-22 manuscript/supp-mat.tex
Manuscript by JJ and TL.

\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\mathrm{fix}}}
\newcommand{\Spop}{\beta}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\Sphy}{S}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\polyDel}{\proba \left[ {\Spop < -1} \right]}
\newcommand{\polyNeutral}{\proba \left[ -1 < \Spop < 1 \right]}
\newcommand{\polyAdv}{\proba \left[ 1 < \Spop  \right]}

@@

    \subsection{Summary table}
    \begin{center}
        \begin{longtable}{|l|l|r|r|r|r|}
            \toprule
            Population                     & Species             & Tajima $\theta_{\pi}$ & $\mathbb{P}_{obs}[1<S]$ & $\mathbb{P} [ 1<\beta ]$ & $\mathbb{P} [ 1<\beta  | 1<S]$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{6}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

@@

    \subsubsection{Proportion of nearly-neutral mutations ($\polyNeutral$)}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Seq0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg-weak_P-Seq0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.pos-weak_P-Seq0.scatter.pdf}
    \end{minipage}

    \subsubsection{Proportion of deleterious mutations ($\polyDel$)}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Sinf0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg_P-Sinf0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg-strong_P-Sinf0.scatter.pdf}
    \end{minipage}

    \subsubsection{Proportion of advantageous mutations ($\polyAdv$)}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Ssup0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.pos_P-Ssup0.scatter.pdf}
    \end{minipage}

@@

    \section{Site-specific codon models.}
    \input{include-Omega-adaptive}
\end{document}

>>>>

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\mathrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\mathrm{fix}}}

\newcommand{\Sphy}{S}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}

\newcommand{\Spop}{\beta}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}

@@

    In BayesCode (\url{https://github.com/ThibaultLatrille/bayescode}), mutation-selection codon models are obtained by running \textit{mutselomega} for 2000 points of MCMC with the options:
    \begin{scriptsize}
        \begin{verbatim}
        mutselomega ---omegashift 0.0 --ncat 30 -a my_alignment.phy -t my_tree.newick -u 2000 my_genename
        \end{verbatim}
    \end{scriptsize}
    The collection of site-specific fitness profiles ($\UniDimArray{F^{(i)}}, \forall i$) are then obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
        readmutselomega --every 1 --until 2000 --burnin 1000 --ss my_genename
        \end{verbatim}
    \end{scriptsize}
    The gene-specific mutation matrix ($\UniDimArray{\mu}$) is also obtained by running \textit{readmutselomega}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
    \begin{scriptsize}
        \begin{verbatim}
        readmutselomega --every 1 --until 2000 --burnin 1000 --nuc my_genename
        \end{verbatim}
    \end{scriptsize}

    \subsection{Summary table}

    \begin{center}
        \begin{longtable}{|l|l|r|r|r|r|}
            \toprule
            Population                     & Species             & Tajima $\theta$ & $\proba_{obs}[\divAdv]$ & $\PpolyAdv$ & $\proba [ \polyAdv | \divAdv]$ \\
            \midrule
            \endhead
            \midrule
            \multicolumn{6}{r}{{Continued on next page}} \\
            \midrule
            \endfoot

@@

    \begin{itemize}
        \item Tajima $\theta$ is the synonymous diversity.
        \item $\proba_{obs}[\divAdv]$ is the proportion of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) larger than 1.
        \item $\PpolyAdv $ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) larger than 1, estimated for the whole genome.
        \item $\proba [ \polyAdv | \divAdv]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) larger than 1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) larger than 1.
    \end{itemize}

@@

    \subsubsection{Proportion of nearly-neutral mutations ($\polyNeutral$)}

    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Seq0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg-weak_P-Seq0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.pos-weak_P-Seq0.scatter.pdf}
    \end{minipage}

    \begin{itemize}
        \item $\ \proba [ \polyNeutral ]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) between -1 and 1, estimated for the whole genome.
        \item $\ \proba [ \polyNeutral | \divWeakDel]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) between -1 and 1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) between -1 and 0.
        \item $\ \proba [ \polyNeutral | \divWeakAdv ]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) between -1 and 1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) between 0 and 1.
    \end{itemize}

    \subsubsection{Proportion of deleterious mutations ($\polyDel$)}

    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Sinf0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg_P-Sinf0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.neg-strong_P-Sinf0.scatter.pdf}
    \end{minipage}

    \begin{itemize}
        \item $\ \proba [ \polyDel ]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) lower than -1, estimated for the whole genome.
        \item $\ \proba [ \polyDel | \divDel]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) lower than -1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) between -3 and -3.
        \item $\ \proba [ \polyDel | \divStrongDel ]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) lower than -1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) lower than -3.
    \end{itemize}

    \subsubsection{Proportion of advantageous mutations ($\polyAdv$)}

    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.all_P-Ssup0.scatter.pdf}
    \end{minipage}
    \begin{minipage}{0.32\linewidth}
        \includegraphics[width=\linewidth, page=1]{../experiments/polyDFE-mC-non-adaptive/regression/results.watterson.pos_P-Ssup0.scatter.pdf}
    \end{minipage}

    \begin{itemize}
        \item $\ \proba [ \polyAdv  ]$ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) greater than 1, estimated for the whole genome.
        \item $\ \proba [ \polyAdv | \divAdv] $ is the proportion of DFE with a selection coefficient at the population-genetic scale ($\Spop$) greater than 1, estimated of SNPs observed with a selection coefficient at the phylogenetic scale ($\Sphy$) greater than 1.
    \end{itemize}

@@

    \section{Site-specific $\omega$-based codon models.}
    The $61$-by-$61$ codon substitution matrix ($\UniDimArray{q^{(i)}}$) at site $i$ is defined entirely by the mutation matrix ($\UniDimArray{\mu}$), $\omega^{(i)}$ and the genetic code:
    \begin{equation}
        \begin{dcases}
            q_{a \mapsto b}^{(i)} & = 0 \text{ if codons $a$ and $b$ are more than one mutation away,} \\
            q_{a \mapsto b}^{(i)} & = \mu_{a \mapsto b} \text{ if codons $a$ and $b$ are synonymous,} \\
            q_{a \mapsto b}^{(i)} & = \omega^{(i)} \mu_{a \mapsto b} \text{ if codons $a$ and $b$ are non-synonymous}.
        \end{dcases}
    \end{equation}
    By definition of the instantaneous rate matrix, the sum of the entries in each row of the codon substitution rate matrix $\UniDimArray{q}$ is equal to $0$, giving the diagonal entries:
    \begin{equation}
        q_{a \mapsto b}^{(i)}= - \sum\limits_{b \neq a, b = 1}^{61} q_{a \mapsto b}^{(i)}.
    \end{equation}

    In BayesCode (\url{https://github.com/ThibaultLatrille/bayescode}), $\omega$-based site-specific codon models are obtained by running \textit{mutselomega} with the options:
    \begin{scriptsize}
        \begin{verbatim}
        mutselomega --omegashift 0.0 --freeomega --omegancat 30 --flatfitness -a my_alignment.phy -t my_tree.newick -u 2000 my_genename
        \end{verbatim}
    \end{scriptsize}
    The mean value of $\omega$ per site is then obtained by running \textit{readmutselomega} with the options:
    \begin{scriptsize}
        \begin{verbatim}
        readmutselomega --every 1 --until 2000 --burnin 1000 -c 0.025 my_genename
        \end{verbatim}
    \end{scriptsize}

    \input{include-Omega-adaptive}
\end{document}
